Manuscript draft

### 2024-03-28 docs/manuscript.qmd
some updates to methods

---
title: "Manuscript draft"
format:
  html:
    embed-resources: true
editor: visual
date: last-modified
author: Kaija Gahm, [others]
bibliography: references.bib
csl: the-american-naturalist.csl
---

@@

-   Multiple spatial behaviors → Social centrality @albery2021

-   Exploration → Weak associations with more conspecifics, more movement between flocks. @aplin2013

-   Slow-exploring → More stable relationships @aplin2013

-   Personality → Social structure @aplin2013

-   Stress → Weaker parental bonds, in captivity @boogert2014

-   Stress → Foraging with more others, less choosy, more central, in captivity @boogert2014

-   Stress → Foraging with more others, less choosy, more central, in the wild @brandl2019

-   More exploratory → higher degree, in finches @moyers2018

-   Quick explorers → More interactions @moyers2018

-   Slower explorers → less central ("Yet, slower explorers overall did not travel shorter distances than faster explorers, indicating that a less central social network position was not merely a consequence of lower activity.") @snijders2014

-   Females → higher strength, more preferred partners (only during lactation) @wey2013

-   Females → more stable partnerships @wey2013

@@

-   Definition of "spatial phenotype" @webber2023

-   Definition of "spatial behaviour" @albery2021

@@

Griffon vultures interact in several social situations, each with a different biological relevance. In particular, griffons rely on social information to locate food resources. Interactions in flight (i.e. when individuals fly within each other's perceptual distance) provide an opportunity for local enhancement of carcass resources, or following to a carcass site. Therefore, they are important for information transfer [@harel2017], finding locations of thermal uplift [@sassi2023], and feeding sites [@spiegel2013; @cortés-avizanda2014]. Vultures also obtain social information about food resources at communal roosting sites. Roost sites can vary in size and density, with many vultures roosting together at once \[REF\]. These roosts also serve as breeding colonies during the breeding season (approximately December through May). Finally, vultures interact frequently at or around carcasses while eating. The free-living griffon vultures in the southern Israel population are provisioned at several feeding sites, where livestock carcasses are deposited irregularly to supplement the availability of naturally-occurring carcasses on the landscape. Vultures may co-feed, associate in close proximity near a carcass, or even fight over the food resource.

@@

For this study, we began with data between September 1, 2020 and September 15, 2023 from the \[ornitela\] and \[inpa\] datasets, which we downloaded from Movebank [@kays2022]. For each vulture, we removed any periods of time spent in the wildlife hospital (e.g. while recovering from an illness, injury, or poisoning), or in captivity prior to wild release (in the case of captive-born birds). We cleaned the data by filtering out GPS fixes deemed unreliable (e.g. too few satellites for triangulation, nonsensical heading readings, impossible jumps from previous fixes), and we removed any time that vultures spent in capture cages during the tagging season in the fall. Altogether, approximately \[PERCENT\] of fixes were removed as compared to the raw data. Next, we removed any vultures that spent fewer than 30 days, in total, within a focal region encompassing the general vicinity of Israel (to exclude, for instance, birds that were briefly detected in the area before dispersing far away). For vultures detected in the area for at least 30 days during the study period, we subsequently removed any fixes for those individuals that occurred outside of the same focal region, in order to restrict our analyses to movements that occurred near the main population locus and ensure that most social interactions would be with other tagged individuals. (We accounted for missing time periods, including removed data, by calculating association indices in the social networks).

@@

After this cleaning, the data used for social network analysis contained \[NUMBER\] of unique vultures in total. Table \[NUMBER\] shows the number of individuals included in the social networks for each season.

@@

### Further cleaning for movement analysis

To characterize individual movement behavior and spatial phenotype, we cleaned the data further. We restricted movement points to the period between sunrise and sunset (calculated for each day with respect to a point at the centroid of Israel using the R package suncalc [@thieurmel2022]. We restricted movement calculations to a set of focal individuals with activity regions centered in the Negev region of Israel, the site of most of our tagging effort. For each vulture, we also removed days with too few GPS fixes. Fixes were ideally collected every 10 minutes, but sometimes they were less frequent due to bad cellular reception in the cliff areas where vultures often sit or due to low battery charge. For each individual, only days with 10 or more GPS fixes per day were retained, unless the battery charge remained above 50% for the entire day. In those cases, the low number of points was attributed to the vultures spending most of their time sitting on a cliff (with limited reception) without much movement, rather than to actual missing data due to low battery. Since these sedentary days represent an important component of vulture movement behavior, we retained this information (Orr Spiegel, pers. obs.).

After removing days with too few fixes per day, we removed any vultures that did not have at least 30 days of valid movement data remaining in a given season, to ensure that movement estimates and averages were based on a high enough sample size of days to capture the range of the vulture's behavior. Since this was done on a per-season basis, individuals that were removed from one season's analysis due to not having enough valid days could still have movement measures calculated for a different season. Finally, we downsampled the movement data to a 10-minute interval to remove high-frequency fixes.

\[Geography\]

For each season, we calculated movement measures only for individuals tracked for at least 30 days (after removing low-point days as described above). We calculated flight altitude by extracting ground elevation information from elevation rasters obtained from the `elevatr` R package [@hollister2022], at a zoom level of \[level--more details about resolution?\] and subtracting those numbers from the vultures' tag-mounted altimeter readings. Finally, since some tags were occasionally set to a much higher fix rate (e.g. every 1 minute) for short periods of time, we downsampled the data to a 10 minute fix rate to make sure that all individuals had comparable fix rates.

## Movement calculations

For each focal individual, we characterized a pattern-based spatial phenotype (Webber et al. 2023) by calculating a variety of movement and space use measures over the course of each of the eight seasons. We calculated home range as the 95% AKDE of a vulture's GPS fixes in a season \@

@@

@@

We used the R package `igraph` [@csardi2005] to calculate each individual's degree and strength for each social situation in each season. Degree is defined as the number of other individuals that a vulture is connected to (i.e. ever interacted with). Strength measures how much an individual interacts, and is calculated as the sum of the weights of all edges that the focal vulture is involved in.

@@

We modeled the relationship between the spatial phenotype variables and the network measures using linear mixed-effects regression models, fitted using the lme4 and lmerTest packages [@bates2015; @kuznetsova2017]. We fit separate models for each response variable (degree and strength). The predictors included were movement (PC1), exploration (PC2), situation (feeding, flight, roosting), age group (juvenile/subadult, adult), and season (breeding, summer, fall). To account for the different numbers of individuals in the network in each unique season, we included a random effect for the interaction of season and year, e.g. "2022_fall" (8 levels). To account for consistent individual differences in network position or behavior, and because individuals varied in how many seasons they were included in, we also included an individual-level random effect.

@@

For each response variable, we began by fitting a model without any interaction terms to check for multicollinearity between the predictors and check model assumptions, using the performance and DHARMa packages [@lüdecke2021; @hartig2022]. Next, we fitted a model containing two three-way interaction terms (between season, situation, and each of the spatial phenotype variables), all of the included two-way interaction terms, age as a main effect, and the two random effects (individual ID; season by year). Examining the model summaries and p values produced by lmerTest, we then removed non-significant higher-order interaction terms, one by one, stopping when all remaining interaction terms were significant. All main and random effects were retained in the final model.

For each model, we calculated marginal effects using the emmeans package [@lenth2023] and created visualizations using ggeffects and ggplot2 [@wickham2016; @lüdecke2018].

@@

```{r}
#| include: false
source(here("scripts/04.0_mixedModels.R"))
```

```{r}
#| echo: false
#| label: fig-d-space
#| fig-cap: "Relationship between normalized degree and space use"
plot_d_eff_space
```

```{r}
#| echo: false
#| label: fig-d-roost
#| fig-cap: "Relationship between normalized degree and roost diversification"
plot_d_eff_roost
```

```{r}
#| echo: false
#| label: fig-d-space-situ
#| fig-cap: "Relationship between normalized degree and space use, with separate trend lines for each social situation"
plot_d_eff_situ_space
plot_d_emt_situ_space
```

```{r}
#| echo: false
#| label: fig-s-movement
#| fig-cap: "Relationship between normalized strength and movement"
plot_s_eff_movement
```

```{r}
#| echo: false
#| label: fig-d-movement-situ
#| fig-cap: "Relationship between normalized degree and movement, with separate trend lines for each social situation"
plot_d_eff_situ_movement
plot_d_emt_situ_movement
```

```{r}
#| echo: false
#| label: fig-s-roost
#| fig-cap: "Relationship between normalized strength and roost diversification"
plot_s_eff_roost
```

```{r}
#| echo: false
#| label: fig-s-space
#| fig-cap: "Relationship between normalized strength and space use"
plot_s_eff_situ_space
plot_s_emt_situ_space
```

>>>>

---
title: "Manuscript draft"
format:
  html:
    embed-resources: true
editor: visual
date: last-modified
author: Kaija Gahm, [others]
bibliography: references.bib
---

@@

-   Multiple spatial behaviors → Social centrality [@albery2021]

-   Exploration → Weak associations with more conspecifics, more movement between flocks. [@aplin2013]

-   Slow-exploring → More stable relationships [@aplin2013]

-   Personality → Social structure [@aplin2013]

-   Stress → Weaker parental bonds, in captivity [@boogert2014]

-   Stress → Foraging with more others, less choosy, more central, in captivity [@boogert2014]

-   Stress → Foraging with more others, less choosy, more central, in the wild [@brandl2019]

-   More exploratory → higher degree, in finches [@moyers2018]

-   Quick explorers → More interactions [@moyers2018]

-   Slower explorers → less central ("Yet, slower explorers overall did not travel shorter distances than faster explorers, indicating that a less central social network position was not merely a consequence of lower activity.") [@snijders2014]

-   Females → higher strength, more preferred partners (only during lactation) [@wey2013]

-   Females → more stable partnerships [@wey2013]

@@

-   Definition of "spatial phenotype" [@webber2023]

-   Definition of "spatial behaviour" [@albery2021]

@@

Griffon vultures interact in several social situations, each with a different biological relevance. In particular, griffons rely on social information to locate food resources. Interactions in flight (i.e. when individuals fly within each other's perceptual distance) provide an opportunity for local enhancement of carcass resources, or following to a carcass site. Therefore, they are important for information transfer [@harel2017], finding locations of thermal uplift [@sassi2023], and feeding sites [@spiegel2013; @cortés-avizanda2014]. Vultures also obtain social information about food resources at communal roosting sites. Roost sites can vary in size and density, with many vultures roosting together at once. These roosts also serve as breeding colonies during the breeding season (approximately December through May). Finally, vultures interact frequently at or around carcasses while eating. The free-living griffon vultures in the southern Israel population are provisioned at several feeding sites, where livestock carcasses are deposited irregularly to supplement the availability of naturally-occurring carcasses on the landscape. Vultures may co-feed, associate in close proximity near a carcass, or even fight over the food resource.

@@

```{r}
#| include: false
load(here("data/dataPrep/dataPrep_report_stats.Rda"))
dp <- dataPrep_report_stats
cl <- which(dp$dataset == "data_cleaned")
pctremovedcleaned <- round(100*(dp$nrow[[cl-1]] - dp$nrow[[cl]])/(dp$nrow[[cl-1]]), 2)
```

For this study, we began with data between September 1, 2020 and September 15, 2023 from the \[ornitela\] and \[inpa\] datasets, which we downloaded from Movebank [@kays2022]. For each vulture, we removed any periods of time spent in the wildlife hospital (e.g. while recovering from an illness, injury, or poisoning), or in captivity prior to wild release (in the case of captive-born birds). We cleaned the data by filtering out GPS fixes deemed unreliable (e.g. too few satellites for triangulation, nonsensical heading readings, impossible jumps from previous fixes), and we removed any time that vultures spent in capture cages during the tagging season in the fall. Altogether, approximately `r pctremovedcleaned`% of fixes were removed as compared to the raw data. Next, we removed any vultures that spent fewer than 30 days, in total, within a focal region encompassing the general vicinity of Israel (in order to exclude, for instance, birds that were briefly detected in the area before dispersing far away). For vultures detected in the area for at least 30 days during the study period, we subsequently removed any fixes for those individuals that occurred outside of the same focal region, in order to restrict our analyses to movements that occurred near the main population locus and ensure that most social interactions would be with other tagged individuals. (We accounted for missing time periods, including removed data, by calculating association indices in the social networks).

@@

```{r}
#| include: false
soc <- which(dp$dataset == "downsampled_10min_forSocial")
vulturessoc <- unique(unlist(dp$vultures[[soc]]))
nvulturessoc <- length(vulturessoc)
nvulturessoc_szns <- dp$nvultures[[soc]]
```

After this cleaning, the data used for social network analysis contained `r nvulturessoc` unique vultures in total (range: `r min(nvulturessoc_szns)`-`r max(nvulturessoc_szns)`; mean = `r round(mean(nvulturessoc_szns), 2)` vultures per season, sd = `r round(sd(nvulturessoc_szns), 2)` vultures per season).

@@

#### Further cleaning for movement analysis

To characterize individual movement behavior and spatial phenotype, we cleaned the data further. We restricted movement points to the period between sunrise and sunset (calculated for each day with respect to a point at the centroid of Israel using the R package {suncalc} [@thieurmel2022]. We restricted movement calculations to a set of focal individuals with activity regions centered in the Negev region of Israel, the site of most of our tagging effort. For each vulture, we also removed days with too few GPS fixes. Fixes were ideally collected every 10 minutes, but sometimes they were less frequent due to bad cellular reception in the cliff areas where vultures often sit or due to low battery charge. For each individual, only days with 10 or more GPS fixes per day were retained, unless the battery charge remained above 50% for the entire day. In those cases, the low number of points was attributed to the vultures spending most of their time sitting on a cliff (with limited reception) without much movement, rather than to actual missing data due to low battery. Since these sedentary days represent an important component of vulture movement behavior, we retained this information (Orr Spiegel, pers. obs.).

```{r}
#| include: false
mov <- which(dp$dataset == "downsampled_10min")
vulturesmov <- unique(unlist(dp$vultures[[mov]]))
nvulturesmov <- length(vulturesmov)
nvulturesmov_szns <- dp$nvultures[[mov]]
```

After removing days with too few fixes per day, we removed any vultures that did not have at least 30 days of valid movement data remaining in a given season, to ensure that movement estimates and averages were based on a high enough sample size of days to capture the range of the vulture's behavior. Since this was done on a per-season basis, individuals that were removed from one season's analysis due to not having enough valid days could still have movement measures calculated for a different season. Finally, we downsampled the movement data to a 10-minute interval to remove high-frequency fixes. Ultimately, we calculated movement measures for a total of `r nvulturesmov` **focal individuals** in total (range: `r min(nvulturesmov_szns)`-`r max(nvulturesmov_szns)`; mean = `r round(mean(nvulturesmov_szns), 2)` vultures per season, sd = `r round(sd(nvulturesmov_szns), 2)` vultures per season).

## Measuring spatial phenotype

For each focal individual, we characterized a pattern-based spatial phenotype (Webber et al. 2023) by calculating a variety of movement and space use measures over the course of each of the eight seasons. We used several measures to capture an individual's space use, roosting behavior, and flight movements.

### Space use

We calculated seasonal home ranges and core areas for each focal vulture, using autocorrelated kernel density estimates (AKDEs) via the {ctmm} R package [@fleming2023]. We defined the home range as the 95% AKDE and the core area as the 50% AKDE. To measure differences in how individuals used their home ranges, we calculated the ratio of the seasonal core area size to the seasonal home range size. We defined a general "space use" variable as the first principal component of a PCA containing the home range size and the core area/home range ratio. The "space use" principal component explained \[PERCENT\] of the variation in the data, with individuals low on the principal component having \[CHARACTERISTICS\] and individuals high on the principal component having \[CHARACTERISTICS\].

### Roost diversification

@@

### Movement

We calculated flight altitude by extracting ground elevation information from elevation rasters obtained from the {elevatr} package [@hollister2022], at a zoom level of \[level--more details about resolution?\] and subtracting those numbers from the vultures' tag-mounted altimeter readings. Finally, since some tags were occasionally set to a much higher fix rate (e.g. every 1 minute) for short periods of time, we downsampled the data to a 10 minute fix rate to make sure that all individuals had comparable fix rates.

@@

We used the R package {igraph} [@csardi2005] to calculate each individual's degree and strength for each social situation in each season. Degree is defined as the number of other individuals that a vulture is connected to (i.e. ever interacted with). Strength measures how much an individual interacts, and is calculated as the sum of the weights of all edges that the focal vulture is involved in.

@@

We modeled the relationship between the spatial phenotype variables and the network measures using linear mixed-effects regression models, fitted using the {lme4} and {lmerTest} packages [@bates2015; @kuznetsova2017]. We fit separate models for each response variable (degree and strength). The predictors included were movement (PC1), exploration (PC2), situation (feeding, flight, roosting), age group (juvenile/subadult, adult), and season (breeding, summer, fall). To account for the different numbers of individuals in the network in each unique season, we included a random effect for the interaction of season and year, e.g. "2022_fall" (8 levels). To account for consistent individual differences in network position or behavior, and because individuals varied in how many seasons they were included in, we also included an individual-level random effect.

@@

For each response variable, we began by fitting a model without any interaction terms to check for multicollinearity between the predictors and check model assumptions, using the {performance} and {DHARMa} packages [@lüdecke2021; @hartig2022]. Next, we fitted a model containing two three-way interaction terms (between season, situation, and each of the spatial phenotype variables), all of the included two-way interaction terms, age as a main effect, and the two random effects (individual ID; season by year). Examining the model summaries and p values produced by lmerTest, we then removed non-significant higher-order interaction terms, one by one, stopping when all remaining interaction terms were significant. All main and random effects were retained in the final model.

For each model, we calculated marginal effects using the {emmeans} package [@lenth2023] and created visualizations using {ggeffects} and {ggplot2} [@wickham2016; @lüdecke2018].

@@

```{r}
#| eval: false
source(here("scripts/04.0_mixedModels.R"))
```

```{r}
#| eval: false
#| label: fig-d-space
#| fig-cap: "Relationship between normalized degree and space use"
plot_d_eff_space
```

```{r}
#| eval: false
#| label: fig-d-roost
#| fig-cap: "Relationship between normalized degree and roost diversification"
plot_d_eff_roost
```

```{r}
#| eval: false
#| label: fig-d-space-situ
#| fig-cap: "Relationship between normalized degree and space use, with separate trend lines for each social situation"
plot_d_eff_situ_space
plot_d_emt_situ_space
```

```{r}
#| eval: false
#| label: fig-s-movement
#| fig-cap: "Relationship between normalized strength and movement"
plot_s_eff_movement
```

```{r}
#| eval: false
#| label: fig-d-movement-situ
#| fig-cap: "Relationship between normalized degree and movement, with separate trend lines for each social situation"
plot_d_eff_situ_movement
plot_d_emt_situ_movement
```

```{r}
#| eval: false
#| label: fig-s-roost
#| fig-cap: "Relationship between normalized strength and roost diversification"
plot_s_eff_roost
```

```{r}
#| eval: false
#| label: fig-s-space
#| fig-cap: "Relationship between normalized strength and space use"
plot_s_eff_situ_space
plot_s_emt_situ_space
```
